Using CopyKAT to idenfity tumor cells in Ewing sarcoma samples

### 2024-05-06 analyses/ewings-cell-type/03-copykat.Rmd
rename notebook and add comments

---
title: "Copy number inference with Ewing sarcoma samples"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
params:
  sample_id: SCPCS000490
  library_id: SCPCL000822
---

@@

@@

# read in ck results 
ck_results_df <- ck_results_files |> 
  purrr::map(readr::read_tsv) |> 
  dplyr::bind_rows(.id = "reference_used")
```

@@

@@

```{r}
caret_df <- cnv_df |> 
  dplyr::filter(copykat.pred != "not.defined") |> 
  dplyr::mutate(copykat = ifelse(
    # use str_detect for test data
    copykat.pred == "diploid", "Normal", "Tumor"
  )) |> 
  split(cnv_df$reference_used)

@@

@@

Now compare to CellAssign and SingleR annotations.

@@

# create jaccard matrices for SingleR and CellAssign compared to tumor/normal 
jaccard_matrices <- celltype_columns |>
  purrr::map(\(name) {
    make_jaccard_matrix(
      cnv_df,
      "copykat.pred",
      name
    )
  }) |> 
  purrr::set_names("SingleR", "CellAssign")
```

@@

>>>>

---
title: "Using CopyKAT to idenfity tumor cells in Ewing sarcoma samples"
author: Ally Hawkins
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
params:
  sample_id: SCPCS000490
  library_id: SCPCL000822
---

@@

This notebook explores using [`CopyKAT`](https://github.com/navinlabcode/copykat) to estimate tumor and normal cells in `r params$sample_id` from SCPCP000015.

`CopyKAT` was run using the `run-copykat.R` script and specifying endothelial cells from both `SingleR` and `CellAssign` as the normal reference. 
This produced results from `CopyKAT` both with and without a normal reference. 
These results are read into this notebook and used to:

- Visualize diploid and anueploid cells on the UMAP. 
- Calculate the confusion matrix comparing manual annotations of tumor cells to using `CopyKAT` to annotate tumor cells. 
- Compare the annotations from `CopyKAT` to cell type annotations from `SingleR` and `CellAssign`.

@@

# read in ck results from both reference types (no_normal and with_normal)
ck_results_df <- ck_results_files |> 
  purrr::map(readr::read_tsv) |> 
  dplyr::bind_rows(.id = "reference_used")
```

@@

### UMAP

Below we prepare and plot a UMAP that shows which cells are classified as diploid, aneuploid, and not defined by `CopyKAT`. 
We show a side by side UMAP with results from running `CopyKAT` both with and without a reference of normal cells.

@@

If we look at the UMAPs shown in `01-marker-gene-tumor-classification.Rmd`, the cells that are labeled aneuploid here are consistent with where we saw marker gene expession for tumor cells. 
It also looks like there is not a visible difference between using a reference of normal cells and not using a reference.

### Confusion Matrix

Below we directly compare the annotations obtained using manual classification of tumor and normal cells to annotating cells with `CopyKAT`. 
To do this, we will calculate the confusion matrix using `caret::confusionMatrix()`.

```{r}
caret_df <- cnv_df |> 
  dplyr::filter(copykat.pred != "not.defined") |> 
  dplyr::mutate(copykat = ifelse(
    # use str_detect for test data
    copykat.pred == "diploid", "Normal", "Tumor"
  )) |> 
  # make tumor the positive class
  dplyr::mutate(
    copykat = forcats::fct_relevel(copykat, "Tumor"),
    marker_gene_classification = forcats::fct_relevel(marker_gene_classification, "Tumor")
  ) |> 
  split(cnv_df$reference_used)

@@

Looking at these results, we see that cells classified as tumor cells using marker genes are mostly overlapping with cells that are aneuploid in CopyKAT, which is what we expect! 
We also see that the Kappa values range are close to 1, indicating almost perfect agreement.

It also looks like there isn't a difference between using the reference and not using the reference. 
The reference Kappa is .1 higher, but I don't think that's enough to argue that you need to use a reference. 
The one thing to note is that using a reference cut the run time for `CopyKAT` in half, so that is one benefit.

We can also calculate the Jaccard similarity index to visualize the amount of cells that have overlapping annotations. 
Given the confusion matrix, I would expect to see good agreement between marker gene and `CopyKAT` annotations.

@@

As expected, it looks like cells classified as tumor cells using marker gene annotation are classified as `aneuploid` with `CopyKAT`.

### Compare CopyKAT to SingleR and CellAssign

Lastly, we will compare the annotations from `CopyKAT` to those obtained using `SingleR` and `CellAssign` by calculating the Jaccard similarity index. 
For this comparison we will use just the annotations from `CopyKAT` with no reference.

@@

# filter to only get annotations from no ref
no_ref_only <- cnv_df |> 
  dplyr::filter(reference_used == "no_ref")

# create jaccard matrices for SingleR and CellAssign compared to aneuploid/diploid 
jaccard_matrices <- celltype_columns |>
  purrr::map(\(name) {
    make_jaccard_matrix(
      no_ref_only,
      "copykat.pred",
      name
    )
  }) |> 
  purrr::set_names("SingleR", "CellAssign")
```

@@

We see that aneuploid cells are mostly muscle cells in both `CellAssign` and `SingleR`. 
In `SingleR` there are some other cell types that were assigned that have a few anueploid cells (chondrocytes, neurons, and fibroblasts).

For diploid cells, we see the most overlap with fibroblasts, endothelial cells, and macrophages. 
These are consistent with the cell types that we see normal cells line up with in `01-marker-gene-tumor-classification.Rmd`.

## Conclusions

- `CopyKAT` seems to do a good job in this Ewing sarcoma sample of identifying aneuploid and diploid cells. 
- The annotations from `CopyKAT` mostly line up with annotations obtained using marker gene annotation. 
- Using a reference of normal cells doesn't change the overall results but does reduce the run time for `CopyKAT`.
